---
title: "Supplementary NatPoKe Figures March2025"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: true
pkgdown:
  as_is: true
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = FALSE)
```

# Introduction

This .html file contains all the additional figures that will compose the Supplementary materials for the NatPoKe article.

```{r, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
library(here)
source(here("src", "libraries.R"))
source(here("src", "speciesResilienceMetricsFigures.R"))
source(here("src", "communityMetricsFigures.R"))
```

# Multi-species abundance over-time

Making use of the .csv file saved as a model output for each species, we will analyze how species populations change over time by combining data from multiple regions.

1.  Create a list, `datasets`, that includes the directories to the .csv files containing the total number of individuals (`TNIND`) data for different continents: Europe, North America, Asia, Africa, and South America.

2.  Then use `do.call("rbind", datasets)` to merge these datasets into a single dataframe called `TNIND_yr`.

3.  Set the two key timepoints: `t_burnin <- 100` marks the end of the burn-in period (a stabilization phase before the actual analysis) and `t_policy <- 110` represents the moment when a specific policy or environmental change occurs, which may impact population trends.

4.  Plot Total Number of Individuals Over Time using `ggplot()`

```{r TNIND plot, fig.dim=c(8,10), warning=FALSE, message=FALSE}
# list directories for all continents output folder
datasets <- list(TNIND_europe, TNIND_northamerica, TNIND_asia, TNIND_africa, TNIND_southamerica)

# bind all df together
TNIND_yr <- do.call("rbind", datasets)

# define key timespoints
t_burnin <- 100
t_policy <- 110

# Custom labeller to show biome name only once per section
custom_labeller <- function(labels) {
  labels$biome <- ifelse(duplicated(labels$biome), "", labels$biome)
  labels
}

TNIND_per_year <- ggplot(data = TNIND_yr, aes(x = timestep, y = TNIND, group = species)) + 
  geom_line() + 
  facet_wrap(biome+region~ species, scales = "free_y", ncol = 4) +
  labs(y = "Total number of individuals") +
  theme_minimal() +
  geom_vline(xintercept = t_policy, linetype = "dotted", color = "black", size = 0.8)  # add line at time of disturbance

TNIND_per_year
```

# Species richness


STILL NEED TO GET TO THIS - NOT SURE IT MAKES ANY SENSE TO HAVE A PLOT OVER TIME. DID ANDRÉ MEAN A SPATIALLY EXPLICIT MAP??


# Model Validation multi-species


# Multispecies model validation

TEST IF FUNCTION CAN REALLY GO THROUGH MULTIPLE DIRECTORIES TO FIND ALL TARGET SPECIES RASTERS.
I THINK IT WORKS BUT I STILL HAVEN'T TESTED!!


Model validation here relies on a previously set user defined function that compares estimated values from our model simulation with an external data source, which in this case came from Santini et al. 2022.

The `validateModel1.2()` function evaluates the performance of our model abundance estimates. It facilitates a direct comparison between model-predicted densities and independent density estimates, providing insights into the model's accuracy. The function:

1.  prepares the independent density data, filtering it to match the target species and formatting it

2.  then it goes through user-specified directories, searching for and reading raster files for species abundance for the specified year. These rasters are then stacked and transformed into a data frame, enabling cell-by-cell analysis. (it also prints a message saying which raster was used for which species) and warns if some are missing.

3.  subsequently, the function calculates the mean predicted abundance for each cell, effectively summarizing the model's output. This information is then joined with the species-specific modeling resolution data, allowing for the calculation of estimated densities. Finally, the function returns a list containing both the processed independent density estimates and the calculated model-predicted densities, facilitating the comparison and validation of the model's performance.

```{r model validation function, warning=FALSE, message=FALSE}
validateModel1.2 <- function(targetspecies, independentDensity, dirouts, spData, validationYear) {
  # Compares mean density estimated by model per cell with
  # predicted density from independent model extract predicted abundance
  # and join with observed abundance
  # THIS FUNCTION WAS ADAPTED FROM ANDRÉS CODE TO WORK THROUGH MULTIPLE DIRECTORIES (dirouts).
  
  ## Species density estimates by an INDEPENDENT SOURCE (akin to observed density)
  independentDensity <- independentDensity %>%
    dplyr::filter(Species %in% targetspecies) %>%
    dplyr::select(Species, lw95, lw75, PredMd, up75, up95) %>%
    mutate(
      lw95 = as.numeric(lw95),
      lw75 = as.numeric(lw75),
      PredMd = as.numeric(PredMd), # Predicted population density (individuals/km2)
      up75 = as.numeric(up75),
      up95 = as.numeric(up95)
    ) %>%
    rename(
      species = Species,
      meanDensity = PredMd
    )
  
  ## Species density estimated by METARANGE from multiple directories
  abundance_files <- list()
  for (target_sps in targetspecies) {
    all_files <- character()
    for (dirout in dirouts) { #Iterate through each directory
      files <- list.files(
        path = dirout,
        pattern = paste0(validationYear, "_", target_sps, "_abundance\\.tif$"),
        full.names = TRUE
      )
      all_files <- c(all_files, files)
    }
    abundance_files[[target_sps]] <- all_files
    if(length(all_files) > 0){
      message(paste("Used rasters for", target_sps, ":", paste(basename(all_files), collapse = ", ")))
    } else {
      warning(paste("No rasters found for", target_sps, "in the given directories."))
    }
  }
  
  abundance_rasters <- lapply(abundance_files, function(files) {
    lapply(files, terra::rast)
  })
  
  abundance_stack_list <- lapply(abundance_rasters, function(raster_list){
    if(length(raster_list) > 0){
      terra::rast(unlist(raster_list))
    } else {
      NULL
    }
  })
  
  # convert raster stack to df
  species_df <- lapply(names(abundance_stack_list), function(sps_name){
    stack <- abundance_stack_list[[sps_name]]
    if(!is.null(stack)){
      lapply(1:terra::nlyr(stack), function(i){
        as.data.frame(stack[[i]], xy = TRUE) %>%
          mutate(species = sps_name) %>%
          rename(abundance = 3)
      }) %>% bind_rows()
    } else {
      NULL
    }
  }) %>% bind_rows()
  
  # format raster's dataframe for validation
  predicted <- species_df %>%
    dplyr::filter(species %in% targetspecies) %>%
    dplyr::group_by(species, x, y) %>%
    dplyr::summarise(
      meanNInd = mean(abundance, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    as.data.frame()
  
  spData2 <- spData %>%
    dplyr::select(Species, ModellingRes) %>%
    rename(species = Species) %>%
    mutate(ModellingRes = ifelse(ModellingRes == unique(ModellingRes)[1], unique(ModellingRes)[1], unique(ModellingRes)[1])) %>% #modified to take the first unique value of ModellingRes
    as.data.frame()
  
  estimatedDensityJoin <- dplyr::inner_join(predicted, spData2, by = "species") %>%
    mutate(estimatedDensity = meanNInd / ModellingRes)
  
  ## compare observed with predicted density
  result_list <- list(independentDensity, estimatedDensityJoin)
  names(result_list) <- c("independentDensity", "estimatedDensity")
  return(result_list)
}
```

```{r multi-species validation, fig.dim=c(8,6), warning=FALSE, message=FALSE}
#load external data
# (2) independentDensity
santini2022 <- read_excel("C:/Users/User/OneDrive - Universidade de Lisboa (1)/ANDRE/SRIT_ANDRE/external_data/geb13476-sup-0002-tables1.xls") %>% 
  # santini's dataframe has species names with spaces but metaRange does not like spaces
  # remove spaces again
  mutate(Species = str_replace_all(Species, " ", ""))

# (4) spData
spData <- read.csv(file.path(dirinput, "metaRangeSpeciesDataframe.csv"))


# applying the function
validation_lists <- validateModel1.2(targetspecies = target_species,
                       independentDensity = santini2022,
                       dirouts = dirout,
                       spData = spData,
                       validationYear = 101)
# plotting the results
ggplot(validation_lists$independentDensity, aes(x = "", y = meanDensity)) +
  geom_boxplot(aes(ymin = lw95, lower = lw75, middle = meanDensity, upper = up75, ymax = up95), stat = "identity") +
  geom_point(data = validation_lists$estimatedDensity, aes(x = "", y = estimatedDensity), color = "red", position = position_jitter(width = 0.2), size = 1) +
  facet_wrap(~ species, scales = "free_y") + 
  ylab("Independent density estimate (individuals/km2)") +
  xlab(" ") +
  ggtitle("Model validation - estimated densities in red") + 
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```


# Sensitivity Analysis

NÃO QUERIA NADA VOLTAR AQUI MAS OK..