---
title: "MetaRange Multispecies Analysis"
author: "Inês Silva"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: true
pkgdown:
  as_is: true
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE)
```

# Introduction

The metaRange model can be run with for multiple species at the same time. Here we demonstrate how to perform a multispecies analysis using the model's outputs.

# Proportion of abundance change across species

To identify and visualize areas where species abundance varies the most, we calculate the proportional change in abundance per grid cell using the following formula:

$Prop.\ Abundance\ Change\ = \frac{Abundance\ t = 125\  -\  abundance\ t = 101} {abundance\ t = 101}$

To achieve this, for each species we:

1.  Import the abundance raster for timestep 101 (after the burn-in phase)

2.  Import the abundance raster for timestep 125 (final simulation timestep)

3.  Compute the proportion of abundance change using the raster files

    Then, to obtain to obtain a multispecies overview of the landscapes

4.  Stack all species' proportion of abundance chnage across species

5.  Compute the mean proportion of abundance change across species

6.  Finally to visualize the results using `ggplot()` we convert the resulting raster into a dataframe

This way we can provide an insight into how species abundance changes across the landscape, helping to identify critical areas of gain or loss of abundance for the taxa.

```{r}
# list for prop abundance change rasters for all sps
prop_abund_change <- list()
# go through each species
for (target_sps in species_names) {
  # import abundance raster after the burn-in for sps
  sps101 <- rast(list.files(path = dirout,
            pattern = paste0(".*101_", target_sps, "_abundance\\.tif"),
            full.names = TRUE))
  # import final abundance raster for sps
  sps125 <- rast(list.files(path = dirout,
                       pattern = paste0(".*125_", target_sps, "_abundance\\.tif"),
                       full.names = TRUE))
  
  # calculate the prop abundance change raster for each sps
  prop_abund_change[[target_sps]] <- (sps125 - sps101) / sps101
}

# Step 2 - Average across all species to get proportion of abund change for MAMMALS
# stack resampled prop abund change rasters for all sps
all_props_stack <- terra::rast(unlist(prop_abund_change))
# average across all layers (species)
mammals_propAbundChange <- terra::mean(all_props_stack, na.rm = TRUE)
# convert to df
mammals_propAbundChange_df <- as.data.frame(mammals_propAbundChange, xy = TRUE)

# Step 3 - Plot mean propotion of abundance change
mammals_abundChange <- ggplot() +
  geom_raster(data = mammals_propAbundChange_df, aes(x = x, y = y, fill = mean)) +
  scale_fill_gradientn(colors = brewer.pal(11, "RdBu"), na.value = "transparent") +
  labs(x = "Latitude", y = "Longitude", fill = "Proportion of\nAbundance\nChange") +
  theme_minimal() +
  theme(legend.title = element_text(size = 10, face = "bold",vjust = 0.9),
        legend.position = c(0.9, 0.2),
        # remove background and line from legend box
        legend.box.background = element_rect(fill = NA, color = NA),
        # remove grid lines
        panel.background = element_blank(),
        panel.grid.minor = element_blank(),
        # remove coordinates and axis labels
        axis.title = element_blank())

# visualise the results
mammals_abundChange
```

# Dispersal change across species

One of the outputs of the metaRange model is a dispersal change raster, which represents the net change in the number of individuals in each grid cell during a given timestep. By averaging dispersal change across all species and timesteps, we can identify key areas of high influx (where individuals are arriving in greater numbers) and high departure (where individuals are leaving the most). The steps bellow explain how we can visualize these results for the mammals in Sweden.

For each species, we:

1.  identify and import all the dispersal change rasters for timestep 101 (first step after the burn-in period) and 125 (last timestep)

2.  Store the rasters within a list

3.  Stack all rasters per species

4.  Compute the mean for each cell across all timesteps

5.  Store the averaged raster per species in a new list

    Then to get an overview for all mammals,

6.  Stack the averaged dispersal change rasters for all species

7.  Compute the overall mean dispersal change

8.  Finally to visualize the results using `ggplot()` we convert the resulting raster into a dataframe

```{r}
# list for dispersal change raster for all timesteps & species
dispChange_list <- list()
# list for dispersal change rasters averaged per species
averaged_dispChange_rasters <- list()
# go through each species
for (target_sps in species_names) {
  
  # Step 1 - Find dispersal change rasters for all species
  
  dispChange_files <- list.files(path = dirout,
                                # import only raster for timesteps 101 to 125
                                pattern = paste0(".*(10[1-9]|11[0-9]|12[0-5])_", target_sps, "_dispersal_change\\.tif"),
                                full.names = TRUE)
  if (length(dispChange_files) > 0) {
    # read all dispersal change rasters for the species
    dispChange_rasters <- lapply(dispChange_files, rast)
    # store all rasters into a list
    dispChange_list[[target_sps]] <- dispChange_rasters
  } else {
    # print a warning if no rasters are found a target species
    dispChange_list[[target_sps]] <- NULL
    print(paste("No abundance files found for", target_sps))
  }
  
  # Step 2 - Average rasters per species
  
  # stack species rasters
  dispChange_stack <- terra::rast(dispChange_list[[target_sps]])
  # calculate the mean across all timesteps (layers)
  averaged_dispChange_raster <- terra::mean(dispChange_stack, na.rm = TRUE)
  # store the averaged raster in a list
  averaged_dispChange_rasters[[target_sps]] <- averaged_dispChange_raster
}

# Step 3 - Average all resampled rasters into one for mammals
dispChange_mammals_stack <- terra::rast(unlist(averaged_dispChange_rasters))
mammals_meanDispChange <- terra::mean(dispChange_mammals_stack, na.rm = TRUE)
# transform into dataframe
mammals_meanDispChange_df <- as.data.frame(mammals_meanDispChange, xy = TRUE)

# Step 4 - Plot mean abundance over time
mammals_dispChange <- ggplot() +
  geom_raster(data = mammals_meanDispChange_df, aes(x = x, y = y, fill = mean)) +
  scale_fill_gradient2(low = "chartreuse3", mid = "white", high = "firebrick3", midpoint = 0, name = "Mean Dispersal\nChange Over\nTime (nº of indiv.)") +
  theme_minimal() +
  theme(legend.title = element_text(size = 10, face = "bold",vjust = 0.9),
        legend.position = c(0.9, 0.2),
        # remove background and line from legend box
        legend.box.background = element_rect(fill = NA, color = NA),
        # remove grid lines
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        # remove coordinates and axis labels
        axis.title = element_blank(),
        axis.text = element_blank())
```
